Redshift Evolution of the Underlying Type Ia Supernova Stretch Distribution

### 2020-12-16 Nicolas.tex
Greg comments included and treated.

%\newcommand{\nn}[1]{\textcolor[rgb]{0.25, 0.50, 0}{#1}}
\newcommand{\nn}[1]{\textcolor[rgb]{0.4, 0.8, 0}{#1}}
\newcommand{\mr}[1]{\textcolor[rgb]{0.5, 0.20, 0.3}{#1}}
\newcommand{\yc}[1]{\textcolor[RGB]{217, 22, 102}{#1}}
\newcommand{\prob}[2]{\mathcal{P}\left( #1 \mid #2\right)}

@@

\author{
    N. Nicolas \thanks{n.nicolas@ipnl.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ipnl.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and J. Nordin\inst{4}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{2,5}
    \orcid{0000-0002-3321-1432}
}

@@

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring typing
and host redshift, the selection effects of magnitude-limited surveys should be
negligible below a given redshift at which even the faintest SNe~Ia can be
observed. \textbf{In practice such spectroscopic following has not always been
the norm as discussed below.} In contrast, targeted surveys have highly
complex selection functions and will be discarded from our analysis.
Fortunately, modern SN cosmology samples such as the Pantheon one are now
dominated by magnitude-limited surveys.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
sample, taking the farthest limits possible. But as this solution might be
over simplistic to create a complete sample, e.g., for it ignores follow-up
efficiency, we also consider another set of cuts to define a conservative
sample. This latter is smaller and therefore less statistically
constraining, but also inherently less prone to selection effects. If the
redshift drift is still significant in the conservative sample, it would be
even more significant in a carefully-tailored selection-free sample. The
goal of this study is to build the first implementation of a SN Ia
properties drift modeling before refining it on a more complex, yet more
accurate, follow-up analysis accounting at the same time for precise selection
modeling using of the SNANA package \citep{SNANA}.}

@@

Similarly, PS1 observed SNe~Ia in the range $0.2<z<0.4$, their $g$-band
$5\sigma$ depth is 23.1~mag \citep{rest2014}, which yields to $z_{\lim}=0.31$ in
agreement with, e.g., Fig.~6 of \cite{scolnic2018a}. \textbf{If we were to be
conservative, this figure would also suggest of a more stringent
$z_{\lim}=0.27$ cut; hence, we will use $0.31$ and $0.27$ for our fiducial and
conservative cuts, respectively.}

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to a $z_{\lim}=0.24$. However,
the SDSS surveys were more sensitive to limited spectroscopic resources
\textbf{as discussed before}. \citet[][Section~2]{kessler2009} pointed out that
during the first year of SDSS, SNe~Ia with $r<20.5$~mag were favored for
spectroscopic follow-up, corresponding to a redshift cut at $0.15$. For the rest
of the SDSS survey, additional spectroscopic resources were available, and
\cite{kessler2009} and \cite{dilday2008} show a reasonable completeness up to
$z_{\lim}=0.2$. Following these analyses, \textbf{we will use $z_{\lim}=0.20$
and $z_{\lim}=0.15$ for out fiducial and conservative cuts, respectively.}

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits \textbf{are roughly located slightly
before the peak of these histograms}. \textbf{We test in
Section~\ref{ssec:verify} the validity of our intent to build volume-limited
surveys from (mostly) magnitude-limited ones with the aforementioned redshift
limits.}

\begin{table}
    \centering
    \caption{Composition of the SNe~Ia dataset used in this analysis.
    Conservative cuts are indicated in parentheses.}
    \label{tab:sample}
    \begin{tabular}{l c c}
        \hline\hline
        Survey & $z_{\lim}$ & $N_{\mathrm{SN}}$ \\
        \hline
        SNf & -- & 114 \\
        SDSS & 0.20 (0.15) & 167 (82)\\
        PS1 & 0.31 (0.27) & 160 (122)\\
        SNLS & 0.60 (0.55) & 102 (78)\\
        HST & -- & 26 \\
        \hline
        Total & -- & 569 (422) \\
        \hline
    \end{tabular}
\end{table}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \textbf{For this dataset,} the spectroscopic follow-up
was done over a redshift range of $0.02 < z < 0.09$, while the search was much
deeper. These 114 SNfactory SNe~Ia are thus in the volume-limited part of the
survey (Aldering et al., in prep.), and are therefore assumed to be a random
sampling of the underlying SN population. The SNfactory sample is particularly
useful for studying SN property drift, as it enables us to have a large complete
SN~Ia sample at $z<0.1$.

@@

\textbf{To test for the existence of potentially significant leftover selection biases in our sample, we compare the stretch and color distributions of the SNe~Ia in our
sample between datasets of overlapping redshifts. If indeed we have built a
volume-limited sample, the distributions of any SNe~Ia parameter at a given
redshift should not depend of their survey of origin. When considering a
redshift range, this has to be narrow enough such that any drift would be
negligible.}

%\yc{YC: \textit{why do you say that here, since you didn't fine-tune
%the overlaping $z$-range?}} \nn{\textit{NN: Because we still took that into
%account?}} \mr{\textit{MR:Because if there is a strong drift in the redshift
%range, and if the redshift distribution of the sample differ, their respective
%property distribution should differ.}}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in the
redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This corresponds to
the 146 SNe~Ia from SDSS at the highest redshift end and thus the most affected
by potential selection effect leftover (see the corresponding discussion in
section~\ref{ssec:cuts}). In that same redshift range, PS1 has 52 SNe~Ia, which
are in the lowest redshift bins and thus unlikely to have any selection issue.
Fig.~\ref{fig:distrib} presents the stretch and color distribution of both these
surveys, together with $p$-values of Kolmogorov-Smirnov tests (KS-test). These
$p$-values are too high ($>10\%$) to exclude that both samples originate from
the same underlying population, in agreement with the visual impression from
Fig.~\ref{fig:distrib}.}

%The $p$-values are high ($>10\%$), and if one cannot conclude from a KS-test
%that indeed these SNe~Ia are from the same underlying distribution, there is no
%ground to claim they are not.

\textbf{We perform a similar analysis for PS1 and SNLS in the redshift range $0.20 < z <
0.31$ where the same conclusion can be drawn, i.e., there is no sign of
discrepancy in the stretch and color distributions between the low- and high-end
of the SNLS and PS1 datasets, respectively. Nonetheless, the small size of the
SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1) limits the interest of the
test as only strong deviation would be noticeable.  Extending the redshift range
to $0.20 < z < 0.40$ (though we have no PS1 data above 0.3) enables to get 31
SNe~Ia from SNLS in total, yet the stretch and color KS-test $p$-values remains
inconclusive, at 34\% and 4\% respectively.}

@@

\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution predicted by
\cite{childress2014} based on SN rates in galaxies depending on their quenching
time as a function of their stellar mass.

@@

@@

Surprisingly, the sample-based Gaussian asymmetric modeling used by current
implementations of the BBC technique \citep{scolnic2016, kessler2017} has one of
the highest AIC value in our analysis (see Section~\ref{sec:results}). While its
$-2\ln(L)$ is the smallest of all redshift-independent models (but still $-11.5$
worse than the reference Base model), it is strongly penalized for requiring
15~free parameters ($\mu_0, \sigma_{\pm}$~for each of the 5~samples of the
analysis). Hence, its $\Delta\mathrm{AIC}<-20$, which could be interpreted as a
probability $p=2\times 10^{-7}$ of being an as good representation of the data
as the Base model.

Remark that, when comparing models adjusted on individual subsamples rather than
globally, the Bayesian Information Criterion ($\mathrm{BIC} = -2\ln(L) +
k\ln(n)$, with $n$ the number of data points) might be better suited than AIC,
since it explicitly accounts for the fact that each subsample is fitted
separately: the sample-based model BIC is rightfully the sum of the BIC for each
sample. Doing so, we find $\Delta\mathrm{BIC}=-48$, again refuting the
sample-based asymmetric Gaussian model as being as pertinent as the Base model.

We report in Table~\ref{tab:bbc} \textbf{our determination of} the samples'
$\mu_0$ and $\sigma_{\pm}$ \textbf{when implementing an asymmetric Gaussian
model}, adjusted on the nominally selection-free samples using our fiducial cuts
(see Section~\ref{sec:sample}). We find our results in close agreement with
\cite{scolnic2016} for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who
derived these model parameters using the full BBC formalism, using numerous
simulations to model the selection effects (see details e.g., Section~3 of
\citealt{kessler2017}). The agreement between our fit of the asymmetric
Gaussians on the supposedly selection-free part of the samples and the results
derived using the BBC formalism supports our approach to get a sample with
negligible selection effects. If we were to use \cite{scolnic2016} and
\cite{scolnic2018a} best fit values of the $\mu_0, \sigma_{\pm}$ asymmetric
parameters for SNLS, SDSS and PS1, respectively, the $\Delta$AIC between our
Base drifting model and the BBC modeling would go even deeper from $-32$ to
$-47$. We further discuss the consequence of this result for cosmology in
Section~\ref{sec:discussion}.

@@

We also performed tests allowing the high-stretch mode of the old population to
differ from the young population mode, hence adding two degrees of freedom. The
corresponding fit is not significantly better, with a $\Delta$AIC of $-0.4$.
This strengthens the fact that the young and old populations indeed appear to
share the same underlying high-stretch mode. Furthermore, one might wonder
whether a low-stretch mode might also exist in the young-population, see
Fig.~\ref{fig:stretchlssfr}. We tested that by allowing this population to also
be bi-modal, finding the amplitude of this low-stretch mode to be compatible
with~0 in this young population ($<2\%$). More generally, this raises the
question of inaccurate tracing of age by a given environmental tracer (here the
LsSFR). A dedicated analysis will be presented in Briday et al.\ in prep.

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though Bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it; see e.g., section 1.3 and 2.5 of
\cite{rubin2015}. Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as
Phillips' relation \citep{phillips1993} standardization parameter $\alpha$ is
not redshift dependent (a study behind the scope of this paper, but see e.g.
\citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias at least for the
upper half of their SN redshift distribution. As a consequence, an ill-modeling
of the underlying stretch distribution will bias the SN magnitude derived from
such surveys.

Commonly used Malmquist bias correction techniques, such as the BBC-formalism,
assume per sample asymmetric Gaussian functions for modeling the underlying
stretch and color distributions. Yet, as shown in Section~\ref{sec:results},
such a sample-based distribution is excluded as being as good as our drifting
model. Then, unlike what \citet[][Section~2]{scolnic2016} and
\citet[][Section~5.4]{scolnic2018a} suggested, i.e. that traditional surveys
span limited redshift ranges and that therefore the per-sample approach accounts
for implicit redshift drifts, a direct modeling of the redshift drift is more
appropriate than a sample-based approach. We stress here that, as measurements
of modern surveys try to cover increasingly larger redshift ranges in order to
reduce calibration systematic uncertainties, this sample-based approach is
becoming less valid, notably for PS1, DES and, soon, LSST.

We illustrate in Fig.~\ref{fig:bbc_pdf_ps1} the prediction difference in the
underlying stretch distribution between the per-sample asymmetric modeling and
our Base drifting model for the PS1 sample. Our model is bimodal and the
relative amplitude of each mode depends on the redshift-dependent fraction of
old and young SNe~Ia in the sample: the higher the fraction of old SNe~Ia (at
lower redshift), the higher the amplitude of the old-specific low-stretch mode.
This redshift dependency is shown as blue to red underlying distributions in
Fig.~\ref{fig:bbc_pdf_ps1} for redshift ranges covered by PS1. The observed
$x_1$ histogram follows our modeling defined as the sum of individual underlying
SN-redshift distributions. As expected, the two modeling approaches differ
mostly in the negative part of the SN stretch distribution. The asymmetric
Gaussian distribution goes through the middle of the bimodal distribution,
over-estimating the number of SNe~Ia at $x_1\sim-0.7$ and under-estimating it at
$x_1\sim-1.7$ in comparison to our Base drifting model for typical PS1 SN
redshifts. This means that the SN bias-corrected standardized magnitude
estimated at a redshift affected by selection effect would be biased by an
ill-modeling of the true underlying stretch distribution.

@@

The amplitude of this magnitude bias for cosmology is beyond the scope of this
paper given the complexity of the BBC analysis. It would require a full study
using our Base model (Eq.~\ref{eq:stretchz}) in place of the sample-based
asymmetric modeling as part of the BBC simulations. However, we already
highlight that even if a non-drifting sample-based model could provide
comparable result in the volume-limited part of the various samples, these
models would differ when extrapolating at higher redshifts, precisely where the
underlying distribution will matter for correcting Malmquist biases.

In the era of modern cosmology, where we aim at probing $w_0$ at a sub-percent
level and $w_a$ at the ten-percent precision \citep[e.g.,][]{lsstpaper}, we
stress that correct modeling of potential SN redshift drift should be further
studied and care should be taken when using samples affected by selection
effects.

@@

We have presented a \textbf{first} study of the drift of the underlying SNe~Ia
stretch distribution as a function of redshift. We built a magnitude-limited
SN~Ia sample from the Pantheon dataset \citep[][SDSS, PS1 and
SNLS]{scolnic2018a}, to which we added HST and SNfactory data from
\cite{rigault2018} for the high- and low-redshift bins. We only considered the
SNe that have been discovered in the redshift range of each survey where
selection effects are negligible, so that the observed SNe~Ia stretches are
random sampling of the true underlying distribution. This resulted in a 569
SN~Ia fiducial sample (422 SNe when more conservative cuts were considered).

Following predictions made in \cite{rigault2018}, we introduced a redshift drift
model which depends on the expected fraction of ``young'' and ``old'' SNe~Ia as
a function of redshift, each age population having its own underlying
stretch distribution.

@@

Our conclusions are the following:
\begin{enumerate}
    \item The underlying SN~Ia stretch distribution is significantly redshift
        dependent, as previously suggested by e.g.~\cite{howell2007} \textbf{in
        a way that selection effects alone cannot explain}. This result is
        largely independent of details on each age-population modeling.

    \item Redshift-independent models are indeed excluded as suitable
        descriptions of the data relative to our Base model. This model assumes
        that: (1) the younger population has a unimodal Gaussian stretch
        distribution, while the older population stretch distribution is
        bimodal, one mode being the same as the young one; (2) the evolution of
        the relative fraction of younger and older SNe~Ia follows the prediction
        made in \cite{rigault2018}. This second result strongly supports the
        existence of both young and old SN~Ia populations, in agreement with
        rate studies \cite{mannucci2005, scannapieco2005, sullivan2006,
        aubourg2008}.

    \item Models using survey-based asymmetric Gaussian distributions, as done,
        e.g., in the current implementation of the BBC, are excluded to be good
        descriptions of the data relative to our drifting model. Hence, the
        sample-based approach does not accurately account for redshift drift and
        even less so as survey span increasingly larger redshift ranges. We
        stipulate that, even if extra degrees of freedom might be acceptable
        given the large number of SNe~Ia in cosmological studies, extrapolating
        the SN property distributions from the volume-limited part of a survey
        to its Malmquist-biased magnitude-limited one would still be inaccurate
        because of the redshift evolution.

>>>>

%\newcommand{\nn}[1]{\textcolor[rgb]{0.25, 0.50, 0}{#1}}
\newcommand{\nn}[1]{\textcolor[rgb]{1.0, 0.6, 0.0}{#1}}
\newcommand{\mr}[1]{\textcolor[rgb]{0.5, 0.20, 0.3}{#1}}
\newcommand{\yc}[1]{\textcolor[RGB]{217, 22, 102}{#1}}
\newcommand{\prob}[2]{\mathcal{P}\left( #1 \mid #2\right)}

@@

\author{
    N. Nicolas \thanks{n.nicolas@ip2i.in2p3.fr, equal contribution} \inst{1} 
    \and M. Rigault \thanks{m.rigault@ip2i.in2p3.fr, equal contribution} \inst{2}
    \orcid{0000-0002-8121-2560}
    \and\\ Y. Copin \inst{1}
    \orcid{0000-0002-5317-7518}
    \and R. Graziani \inst{2}
    \and G. Aldering\inst{3}
    \and M. Briday\inst{1}
    \and Y.-L. Kim\inst{1}
    \orcid{0000-0002-1031-0796}
    \and J. Nordin\inst{4}
    \and Saul Perlmutter\inst{3}
    \and M. Smith\inst{2,5}
    \orcid{0000-0002-3321-1432}
}

@@

Assuming sufficient (and unbiased) spectroscopic follow-up for acquiring typing
and host redshift, the selection effects of magnitude-limited surveys should be
negligible below a given redshift at which even the faintest SNe~Ia can be
observed. \textbf{In practice such spectroscopic following has not always been
the norm as discussed below.} In contrast, targeted surveys have highly complex
selection functions and will be discarded from our analysis.  Modern SN
cosmology samples such as the Pantheon one are now dominated by
magnitude-limited surveys \nn{which have easier to model selection functions}.

@@

\textbf{We will thus consider a set of cuts that will define a first fiducial
sample, taking the farthest limits possible. But as this solution might be
over simplistic to create a complete sample, e.g., for it ignores follow-up
efficiency, we also consider another set of cuts to define a conservative
sample. This latter is smaller and therefore less statistically
constraining, but also inherently less prone to selection effects. If the
redshift drift is still significant in the conservative sample, it would be
even more significant in a carefully-tailored selection-free sample. The
goal of this study is to build the first implementation of a SN Ia
properties drift modeling before refining it on a more complex, yet more
accurate, follow-up analysis accounting at the same time for precise selection
modeling using, e.g., the SNANA package \citep{SNANA}.}

@@

Similarly, PS1 observed SNe~Ia in the range $0.2<z<0.4$, their $g$-band
$5\sigma$ depth is 23.1~mag \citep{rest2014}, which yields $z_{\lim}=0.31$, in
agreement with, e.g., Fig.~6 of \cite{scolnic2018a}. \textbf{If we were to be
    conservative, this figure would also suggest of a more stringent
$z_{\lim}=0.27$ cut; hence, we will use $0.31$ and $0.27$ for our fiducial and
conservative cuts, respectively.}

In a similar redshift range, SDSS has a limiting magnitude of 22.5
\citep{dilday2008, sako2008}, which would lead to $z_{\lim}=0.24$. However, the
SDSS surveys \nn{had to contend with limited spectroscopic resources. As
discussed in} \citet[][Section~2]{kessler2009}, during the first year of SDSS,
SNe~Ia with $r<20.5$~mag were favored for spectroscopic follow-up, corresponding
to a redshift cut at $0.15$. For the rest of the SDSS survey, additional
spectroscopic resources were available, and \cite{kessler2009} and
\cite{dilday2008} show a reasonable completeness up to $z_{\lim}=0.2$. Following
these analyses, \textbf{we will use $z_{\lim}=0.20$ and $z_{\lim}=0.15$ for our
fiducial and conservative cuts, respectively.}

The sample selection is summarized in Table~\ref{tab:sample}, and the redshift
distribution of these three surveys is shown in Fig.~\ref{fig:cuts}. As
expected, the selected redshift limits \textbf{are roughly located slightly
    before the peak of these histograms}. \nn{In Section 2.2 we validate that
    these redshift limits are effective for constructing nearly volume-limited
subsamples from samples that were initially more-closely magnitude limited in
their search or spectroscopic follow-up.}

\begin{table}
    \centering
    \caption{Composition of the SNe~Ia dataset used in this analysis.
    Conservative cuts are indicated in parentheses. \nn{The SNf limit is set by
\cite{rigault2018}, see text.}}
    \label{tab:sample}
    \begin{tabular}{l c c}
        \hline\hline
        Survey & $z_{\lim}$ & $N_{\mathrm{SN}}$ \\
        \hline
        SNf & 0.08 & 114 \\
        SDSS & 0.20 (0.15) & 167 (82)\\
        PS1 & 0.31 (0.27) & 160 (122)\\
        SNLS & 0.60 (0.55) & 102 (78)\\
        HST & -- & 26 \\
        \hline
        Total & -- & 569 (422) \\
        \hline
    \end{tabular}
\end{table}

@@

In addition, we use the SNe~Ia from the Nearby Supernova Factory
\citep[SNfactory,][]{aldering2002} published in \cite{rigault2018} and that have
been discovered from non-targeted searches (114 SNe~Ia, see their sections~3
and~4.2.2; SNe~Ia time series are published in \citealt{saunders2020}, see also
\citealt{aldering2020}). \nn{For this dataset, spectroscopic screening was done
for candidates with $r \lesssim 19.5$. Then, redshift cuts were applied when
selecting which SN Ia to follow, giving a redshift range of 0.02 < z < 0.09 ;
further reduced to $<0.08$ in \cite{rigault2018} for extracting local host
properties.} These 114 SNfactory SNe~Ia are thus in the volume-limited part of
the survey (Aldering et al., in prep.), and are therefore assumed to be a random
sampling of the underlying SN population. The SNfactory sample is particularly
useful for studying SN property drift, as it enables us to have a large complete
SN~Ia sample at $z<0.1$.

@@

\textbf{To test for the existence of potentially significant leftover selection
biases in our sample, we compare the stretch and color distributions of the
SNe~Ia in our sample between datasets of overlapping redshifts. If indeed we
have built a volume-limited sample, the distributions of any SNe~Ia
parameter at a given redshift should not depend on their survey of origin. When
considering a redshift range, this has to be narrow enough such that any drift
would be negligible.}

\textbf{The two samples that overlap the most in redshift are PS1 and SDSS in
the redshift range $0.10 < z < 0.20$ (see Fig~\ref{fig:sample}). This
corresponds to the 146 SNe~Ia from SDSS at the highest redshift end and thus
the most \nn{likely to be} affected by selection effects \nn{not eliminated
by our redshift cut} (see the corresponding discussion in
section~\ref{ssec:cuts}). In that same redshift range, PS1 has 52 SNe~Ia,
which are in the lowest redshift bins and thus unlikely to have any
selection issue.  \nn{To identify potential inconsistency between PS1 and
SDSS, Fig.~\ref{fig:distrib} presents} the stretch and color
distribution of both these surveys, together with $p$-values of
Kolmogorov-Smirnov (KS) tests. \nn{These $p$-values are too high
($>10\%$) and do not indicate that these subsamples are inconsistent}, in
agreement with the visual impression from Fig.~\ref{fig:distrib}.}

\textbf{We perform a similar analysis for PS1 and SNLS in the redshift range
$0.20 < z < 0.31$ where the same conclusion can be drawn, i.e., there is no
sign of discrepancy in the stretch and color distributions between the low-
and high-end of the SNLS and PS1 datasets, respectively. Nonetheless, the
small size of the SNLS dataset at $z < 0.31$ (12 SNe~Ia vs. 90 for PS1)
limits the \nn{sensitivity} of the test, as only strong deviation would be
noticeable. Extending the redshift range to $0.20 < z < 0.40$ (though we
have no PS1 data above 0.3) enables to get 31 SNe~Ia from SNLS in total, yet
\nn{the stretch p-values remains high at 34\% showing no sign of inconsistency}}

@@

\begin{equation}
    \label{eq:delta}
    \delta(z) = \left( K^{-1} \times (1+z)^{-2.8} +1 \right)^{-1}
\end{equation}
with $K=0.87$. This model is comparable to the evolution \nn{subsequently}
predicted by \cite{childress2014} based on SN rates in galaxies depending on
their quenching time as a function of their stellar mass.

@@

\nn{This is our Base model.}

@@

Surprisingly, the sample-based Gaussian asymmetric modeling used by current
implementations of the BBC technique \citep{scolnic2016, kessler2017} has one of
the highest AIC value in our analysis (see Section~\ref{sec:results}). While its
$-2\ln(L)$ is the smallest of all redshift-independent models (but still $-11.5$
worse than the reference Base model), it is strongly penalized for requiring
15~free parameters ($\mu_0, \sigma_{\pm}$~for each of the 5~samples of the
analysis). Hence, its $\Delta\mathrm{AIC}<-20$, which could be interpreted as a
probability $p=2\times 10^{-7}$ of being as good a representation of the data
as the Base model.

\nn{Note} that, when comparing models adjusted on individual subsamples rather
than globally, the Bayesian Information Criterion ($\mathrm{BIC} = -2\ln(L) +
k\ln(n)$, with $n$ the number of data points) might be better suited than AIC,
since it explicitly accounts for the fact that each subsample is fitted
separately: the sample-based model BIC is rightfully the sum of the BIC for each
sample. We find $\Delta\mathrm{BIC}=-48$, again refuting the sample-based
asymmetric Gaussian model as being as pertinent as the Base model.

We report in Table~\ref{tab:bbc} \textbf{our determination of} the samples'
$\mu_0$ and $\sigma_{\pm}$ \textbf{when implementing an asymmetric Gaussian
model}, adjusted on the nominally selection-free samples using our fiducial cuts
(see Section~\ref{sec:sample}). We find our results in close agreement with
\cite{scolnic2016} for SNLS and SDSS and with \cite{scolnic2018a} for PS1, who
derived these model parameters using the full BBC formalism, using numerous
simulations to model the selection effects (see details e.g., Section~3 of
\citealt{kessler2017}). The agreement between our fit of the asymmetric
Gaussians on the supposedly selection-free part of the samples and the results
derived using the BBC formalism supports our approach to \nn{construct} a sample
with negligible selection effects. If we were to use \cite{scolnic2016} and
\cite{scolnic2018a} best fit values of the $\mu_0, \sigma_{\pm}$ asymmetric
parameters for SNLS, SDSS and PS1, respectively, the $\Delta$AIC between our
Base drifting model and the BBC modeling would go even deeper from $-32$ to
$-47$. We further discuss the consequence of this result for cosmology in
Section~\ref{sec:discussion}.

@@

We also performed tests allowing the high-stretch mode of the old population to
differ from the young population mode, hence adding two degrees of freedom. The
corresponding fit is not significantly better, with a $\Delta$AIC of $-0.4$.
This \nn{reinforces our assumption} that the young and old populations indeed
appear to share the same underlying high-stretch mode. Furthermore, one might
wonder whether a low-stretch mode might also exist in the young-population, see
Fig.~\ref{fig:stretchlssfr}. We \nn{test for this} by allowing this population
to also be bi-modal, finding the amplitude of \nn{such a} low-stretch mode to be
compatible with~0 in this young population ($<2\%$). More generally, this raises
the question of\nn{how well a given environmental tracer (here LsSFR) traces the
age}.  A dedicated analysis will be presented in Briday et al.\ in prep.

@@

To the best of our knowledge, a SN~Ia stretch redshift drift modeling has never
been explicitly used in cosmological analyses, though Bayesian hierarchy
formalism such as UNITY \citep{rubin2015}, BAHAMAS \citep{shariff2016} or Steve
\citep{hinton2019} can easily allow it; see e.g., section 1.3 and 2.5 of
\cite{rubin2015}. Not doing so is a second order issue for SN cosmology, as it
only affects the way one accounts for Malmquist bias. Indeed, as long as the
Phillips relation \citep{phillips1993} standardization parameter $\alpha$ is not
redshift dependent (a study \nn{beyond} the scope of this paper, but see e.g.
\citealt{scolnic2018a}), the stretch-corrected SNe~Ia magnitudes used for
cosmology are blind to the underlying stretch distribution for complete samples.
However, surveys usually do have significant Malmquist bias for the upper half
of their SN redshift distribution. As a consequence, an ill-modeling of the
underlying stretch distribution will bias the SN magnitudes derived from such
surveys.

Commonly used Malmquist bias correction techniques, such as the BBC-formalism,
assume per sample asymmetric Gaussian functions for modeling the underlying
stretch and color distributions. Yet, as shown in Section~\ref{sec:results},
such a sample-based distribution is excluded as being as good as our drifting
model. Then, unlike what \citet[][Section~2]{scolnic2016} and
\citet[][Section~5.4]{scolnic2018a} suggested, i.e. that traditional surveys
span limited redshift ranges and that therefore the per-sample approach accounts
for implicit redshift drifts, a direct modeling of the redshift drift is more
appropriate than a sample-based approach. We \nn{add} here that, as measurements
of modern surveys try to cover increasingly larger redshift ranges in order to
reduce calibration systematic uncertainties, this sample-based approach
\nn{becomes} less valid, notably for PS1, DES and, soon, LSST.

We illustrate in Fig.~\ref{fig:bbc_pdf_ps1} the prediction difference in the
underlying stretch distribution between the per-sample asymmetric modeling and
our Base drifting model for the PS1 sample. Our model is bimodal and the
relative amplitude of each mode depends on the redshift-dependent fraction of
old and young SNe~Ia in the sample: the higher the fraction of old SNe~Ia (at
lower redshift), the higher the amplitude of the old-specific low-stretch mode.
This redshift dependency is shown as blue to red underlying distributions in
Fig.~\ref{fig:bbc_pdf_ps1} for redshift ranges covered by PS1. The observed
$x_1$ histogram follows \nn{the model we defined using} the sum of individual
underlying SN-redshift distributions. As expected, the two modeling approaches
differ mostly in the negative part of the SN stretch distribution. The
asymmetric Gaussian distribution goes through the middle of the bimodal
distribution, over-estimating the number of SNe~Ia at $x_1\sim-0.7$ and
under-estimating it at $x_1\sim-1.7$ in comparison to our Base drifting model
for typical PS1 SN redshifts. This means that the SN bias-corrected standardized
magnitude estimated at a redshift affected by selection effects would be biased
by an ill-modeling of the true underlying stretch distribution.

@@

\nn{Assessing t}he amplitude of this magnitude bias for cosmology is beyond the
scope of this paper given the complexity of the BBC analysis. It would require a
full study using our Base model (Eq.~\ref{eq:stretchz}) in place of the
sample-based asymmetric modeling as part of the BBC simulations. However, we
already highlight that even if a non-drifting sample-based model could provide
comparable result in the volume-limited part of the various samples, these
models would differ when extrapolating to higher redshifts, precisely where the
underlying distribution will matter for correcting Malmquist biases.

In the era of modern cosmology, where we aim \nn{to measure} $w_0$ at a
sub-percent level and $w_a$ \nn{with} ten-percent precision
\citep[e.g.,][]{lsstpaper}, we stress that correct modeling of potential SN
redshift drift should be further studied and care should be taken when using
samples affected by selection effects.

@@

We have presented \nn{an initial} study of the drift of the underlying SNe~Ia
stretch distribution as a function of redshift. We built a magnitude-limited
SN~Ia sample from the Pantheon dataset \citep[][SDSS, PS1 and
SNLS]{scolnic2018a}, to which we added HST and SNfactory data from
\cite{rigault2018} for the high- and low-redshift bins. We only considered the
SNe that have been discovered in the redshift range of each survey where
selection effects are negligible, so that the observed SNe~Ia stretches are
a random sampling of the true underlying distribution. This resulted in a 569
SN~Ia fiducial sample (422 SNe when more conservative cuts were considered).

Following predictions made in \cite{rigault2018}, we introduced a redshift drift
model which depends on the expected fraction of ``young'' and ``old'' SNe~Ia as
a function of redshift, \nn{with} each age population having its own underlying
stretch distribution.

@@

Our conclusions are the following:
\begin{enumerate}
    \item The underlying SN~Ia stretch distribution is significantly redshift
        dependent, as previously suggested by e.g.~\cite{howell2007}, \textbf{in
        a way that selection effects alone cannot explain}. This result is
        largely independent of details on each age-population model.

    \item Redshift-independent models are \nn{quantitatively} excluded as
        suitable descriptions of the data relative to our Base model. This model
        assumes that: (1) the younger population has a unimodal Gaussian stretch
        distribution, while the older population stretch distribution is
        bimodal, one mode being the same as the young one; (2) the evolution of
        the relative fraction of younger and older SNe~Ia follows the prediction
        made in \cite{rigault2018}. This second result strongly supports the
        existence of both young and old SN~Ia populations, in agreement with
        rate studies \cite{mannucci2005, scannapieco2005, sullivan2006,
        aubourg2008}.

    \item Models using survey-based asymmetric Gaussian distributions, as done,
        e.g., in the current implementation of BBC, are excluded \nn{as a} good
        description of the data relative to our drifting model. Hence, the
        sample-based approach does not accurately account for redshift drift and
        even less so as survey span increasingly larger redshift ranges.  \nn{As
        a result, if} extra degrees of freedom might be acceptable given the
        large number of SNe~Ia in cosmological studies, extrapolating the SN
        property distributions from the volume-limited part of a survey to its
        Malmquist-biased magnitude-limited one would still be inaccurate because
        of the redshift evolution.
